---
title: "Session 5 -- Data manipulation with dplyr"
subtitle: "Instructor notes"
# Wider body. grid is deep-merged with the project block in _quarto.yml,
# so sidebar-width, margin-width and gutter-width are inherited.
# page-layout: full would NOT widen this page: with toc-location: left
# the TOC is #quarto-sidebar, so Quarto picks the column-page-right
# layout, whose body is still capped at $grid-body-width.
grid:
  body-width: 1400px
toc-depth: 4
toc-expand: true
# Appended to the project css rather than replacing it, so fonts.css
# still loads (Quarto treats css as a mergeable list at every level).
css: assets/css/instructor.css
include-in-header: _includes/instructor-head.html
# Keeps these pages out of docs/search.json. `navbar: search: false`
# only hides the search box - the index is written regardless.
search: false
# Rendered at 480x307 and displayed at half that by instructor.css; see
# the Figures block there for which way that trade goes and why.
fig-format: png
fig-width: 5
fig-height: 3.2
---

<!-- GENERATED by _MakeInstructorNotes.R from _instructor_prompt.md.
     Committed, but do not hand-edit below the BODY marker: the next run
     that sees a changed source page overwrites it. To change the prose,
     edit the prompt; to change the layout, edit this script. -->
<!-- instructor-notes: source=session5.qmd md5=c0591ead8f6605ffbda3c7cad64288a4 commit=25471859b6eac7133160cbe89f84a78b0ffdef91 dirty=no model=opus effort=high -->

```{r}
#| label: instructor-setup
#| include: false
# width/cli.width are set project-wide in _quarto.yml and are restated
# here so that this one chunk is the whole story for a generated page -
# otherwise a later edit here silently drops them.
options(width = 80, cli.width = 80,
        tibble.print_min = 2, tibble.print_max = 2,
        # A 32-column tibble spends six lines listing the columns it
        # could not fit. One line saying how many is enough here - the
        # "1,904 x 32" header already gives the shape.
        tibble.max_extra_cols = 1)

# Base data frames have no equivalent of the tibble options, so they need
# a method. It MUST let tibbles through: they are data frames too, so S3
# dispatch would land them here, and head() would throw away the real
# row count that the tibble footer reports. kable/kableExtra/DT all
# return their own classes, so none of them come through this.
registerS3method("knit_print", "data.frame", function(x, ...) {
    if (inherits(x, "tbl_df")) return(knitr::normal_print(x))
    knitr::normal_print(utils::head(x, 2L))
}, envir = asNamespace("knitr"))
```

::: {.instructor-nav}
[Contents](instructor_index.html) &nbsp;·&nbsp; [1](instructor_session1.html) &nbsp;·&nbsp; [2](instructor_session2.html) &nbsp;·&nbsp; [3](instructor_session3.html) &nbsp;·&nbsp; [4](instructor_session4.html) &nbsp;·&nbsp; **5** &nbsp;·&nbsp; [6](instructor_session6.html) &nbsp;·&nbsp; [7](instructor_session7.html) &nbsp;·&nbsp; [8](instructor_session8.html) &nbsp;·&nbsp; [Student page](session5.html)
:::

<!-- BODY -->
::: {.callout-note .objectives title="Learning objectives"}
* Chain operations together into a workflow using pipes
* Learn the key **dplyr** functions for manipulating your data
* Learn about faceting in ggplot2 to split your data into separate categories
  and create a series of sub-plots arranged in a grid
:::

---

# Data manipulation with dplyr

* **dplyr** loads with the tidyverse -- the Swiss army knife for tabular data

```{r}
#| label: library_tidyverse
library(tidyverse)
```

* Today: `filter()`, `mutate()`, `summarise()`, the pipe `|>`, and ggplot2 **faceting**

```{r}
#| label: load_metabric
#| message: false
metabric <- read_csv("data/metabric_clinical_and_expression_data.csv")
metabric
```

---


# dplyr verbs

* dplyr functions are **"verbs"** in a grammar of data manipulation -- all share the same syntax:

* The first argument in each function is a data frame (or tibble)

* Any additional arguments describe what operation to perform on the data frame

* Variable names, i.e. column names, are referred to without using quotes

* The result of an operation is a new data frame

---

# Chaining operations using `|>`

* Recap: `filter()` then `select()` using an intermediate variable

```{r}
#| label: filter_then_select
patients <- filter(metabric,
                   Survival_status == "LIVING" & Survival_time > 120)
patient_details <- select(patients,
                          Patient_ID,
                          Survival_time,
                          Tumour_stage,
                          Nottingham_prognostic_index)
patient_details
```

* Option 1 -- reuse the same name to avoid cluttering the environment

```{r}
#| label: select_overwriting_patients
#| eval: false
patients <- select(patients,
                   Patient_ID,
                   Survival_time,
                   Tumour_stage,
                   Nottingham_prognostic_index)
```

* Option 2 -- **nesting** function calls; fine for two simple steps

```{r}
#| label: nested_functions_intermediate
x <- c(24, 53, 352, 643, 46)
myMean <- mean(x)
sqrt(myMean)
```

```{r}
#| label: nested_functions
sqrt(mean(x))
```

* Same idea with dplyr -- read inside-out, unwieldy, hard to read

```{r}
#| label: nested_dplyr_calls
patients <- select(filter(metabric, Survival_status == "LIVING", Survival_time > 120), Patient_ID, Survival_time)
nrow(patients)
```

::: {.rmdblock}
**The pipe operator `|>`**

* Takes the left-hand side and passes it as the **first argument** to the function on the right

**`x |> f(y)`** is equivalent to **`f(x, y)`**

For example:

`select(starwars, name, height, mass)`

can be rewritten as

`starwars |> select(name, height, mass)`

Chained into a workflow:

`starwars |>`<br/>
&nbsp;&nbsp;&nbsp;&nbsp;`filter(species == "Droid") |>`<br/>
&nbsp;&nbsp;&nbsp;&nbsp;`select(name, height, mass)`

* Built into R -- no package needed; modelled on the Unix pipe `|`
* Called the **native pipe**; needs R 4.1+
* You will meet **`%>%`** (magrittr, loaded with tidyverse) in older code -- read it as `|>`; we use `|>` throughout
:::

```{r}
#| label: pipe_filter_select
patients <- metabric |>
    filter(Survival_status == "LIVING" & Survival_time > 120) |>
    select(Patient_ID, Survival_time, Tumour_stage, Nottingham_prognostic_index)
```

* Why tidyverse functions always take the data frame first and always return a data frame -- so they pipe
* Build a workflow **one step at a time**, checking the output as you go

```{r}
#| label: pipe_single_step
patients <- metabric |>
    filter(Survival_status == "LIVING")
```

```{r}
#| label: pipe_two_filters
patients <- metabric |>
    filter(Survival_status == "LIVING") |>
    filter(Survival_time > 120)
```

```{r}
#| label: pipe_full_chain
patients <- metabric |>
    filter(Survival_status == "LIVING") |>
    filter(Survival_time > 120) |>
    select(Patient_ID, Survival_time, Tumour_stage, Nottingham_prognostic_index)
# print out the result
patients
```

* **Gotcha:** `|>` must be at the **end** of the line, not the start of the next

```
# R considers the following to be 2 separate commands, the first of which ends
# with the first filter operation and runs successfully.
# The second statement is the third and last line, is not a valid command and
# so you'll get an error message
patients <- metabric |>
    filter(Survival_status == "LIVING")
    |> filter(Survival_time > 120)
```

* Same rule as `+` in ggplot2

---

# Modifying data using `mutate()`

* Base R way -- add a new logical column with `$`

```{r}
#| label: add_column_base_r
metabric$Deceased <- metabric$Survival_status == "DECEASED"
```

* dplyr way -- `mutate()`; new column added at the end; **no `metabric$` prefix needed**

```{r}
#| label: mutate_deceased
metabric <- mutate(metabric, Deceased = Survival_status == "DECEASED")
```

* `mutate()` can **overwrite** an existing column -- common for unit conversion
* Tumour size is in mm, NPI formula needs cm

```{r}
#| label: mutate_tumour_size
metabric <- mutate(metabric, Tumour_size = Tumour_size / 10)
```

::: {.rmdblock}
**Nottingham Prognostic Index**

Prognosis after breast cancer surgery, from three pathological criteria.

**`NPI = (0.2 * S) + N + G`**

* **S** is the size of the tumour in centimetres
* **N** is the node status (0 nodes = 1, 1-3 nodes = 2, >3 nodes = 3)
* **G** is the grade of tumour (Grade I = 1, Grade II = 2, Grade III = 3)
:::

* Recalculate NPI and compare with the column already in the data

```{r}
#| label: mutate_npi
metabric <- metabric |>
    mutate(NPI = (0.2 * Tumour_size) +
               Lymph_node_status +
               Neoplasm_histologic_grade)

select(metabric,
       Tumour_size,
       Lymph_node_status,
       Neoplasm_histologic_grade,
       NPI,
       Nottingham_prognostic_index)
```

* Ask the class: spot the discrepancy -- original NPI used the wrong tumour size scaling

```{r}
#| label: scatter_plot_1
ggplot(data = metabric) +
    geom_point(mapping = aes(x = Age_at_diagnosis, y = NPI), na.rm = TRUE)
```

* Reread the data and build one workflow: convert, compute, **`round()`**, `arrange(desc())`, select

```{r}
#| label: mutate_pipeline
#| message: false
read_csv("data/metabric_clinical_and_expression_data.csv") |>
    mutate(Tumour_size = Tumour_size / 10) |>
    mutate(NPI = (0.2 * Tumour_size) +
               Lymph_node_status +
               Neoplasm_histologic_grade) |>
    mutate(Tumour_size = round(Tumour_size, digits = 1)) |>
    mutate(NPI = round(NPI, digits = 1)) |>
    arrange(desc(NPI)) |>
    select(Tumour_size, Lymph_node_status, Neoplasm_histologic_grade, NPI)
```

## Mutating multiple columns

* Two identical rounding steps above -- collapse into one with **`across()`**
* `across()` goes inside `mutate()`: first argument = columns, second = function to apply

```{r}
#| label: mutate_across_columns
metabric |>
    mutate(across(c(Tumour_size, NPI), round)) |>
    select(Patient_ID, Tumour_size, NPI)
```

* Harder case: no ready-made function, or you need to pass extra arguments
* Use an **anonymous function** -- no name, exists only inside the `mutate()` call
* Example: cm to mm in `iris`; there is no "multiply by 10" function

```{r}
#| label: mutate_across_anonymous_function
iris |>
    as_tibble() |>
    mutate(across(Sepal.Length:Petal.Width, ~ .x * 10))
```

* **`~`** marks the anonymous function (R's formula symbol); **`.x`** is the placeholder for each column
* To pass extra arguments to an existing function: `~func(.x, ...)`

```{r}
#| label: mutate_across_with_arguments
metabric |>
    mutate(across(c(Tumour_size, NPI), ~round(.x, digits = 1))) |>
    select(Patient_ID, Tumour_size, NPI)
```

* Inside `across()` you can use the range operator and the same helpers as `select()`

```{r}
#| label: mutate_across_gene_range
metabric |>
    mutate(across(ESR1:MLPH, ~round(.x, digits = 2))) |>
    select(Patient_ID, ESR1:MLPH)
```

* `ends_with()` helper -- convert all `_status` columns to **factors**

```{r}
#| label: mutate_across_ends_with
metabric |>
    mutate(across(ends_with("_status"), as.factor)) |>
    select(Patient_ID, ends_with("_status"))
```

---

# Computing summary values using `summarise()`

* Brief here -- comes into its own with **`group_by()`** next session
* Any function collapsing a vector to a single value works

```{r}
#| label: mean_esr1_base_r
mean(metabric$ESR1)
```

```{r}
#| label: summarise_mean_esr1
summarise(metabric, mean(ESR1))
```

* `summarize()` spelling also works
* Key difference: `mean()` returns a length-1 vector, `summarise()` returns a **data frame** (1 row, 1 column)
* Useful when summarising several columns or using several functions -- and you can **name** the output columns

```{r}
#| label: summarise_named_values
summarise(metabric, ESR1_mean = mean(ESR1), ESR1_sd = sd(ESR1))
```

::: {.rmdblock}
**`summarise()`**

`summarise()` collapses a data frame into a single row by calculating summary
values of one or more of the columns.

Takes any function that maps a vector to a single value:

* Centre: **`mean()`**, **`median()`**
* Spread: **`sd()`**, **`mad()`**
* Range: **`min()`**, **`max()`**, **`quantile()`**
* Position: **`first()`**, **`last()`**
* Count: **`n()`**

`first()`, `last()` and `n()` are only really useful with **`group_by()`**.

**`n()`** is special -- returns the number of observations, takes no column argument.
:::

* Summary functions can span two columns, e.g. a correlation

```{r}
#| label: summarise_correlation
summarise(metabric, correlation = cor(FOXA1, MLPH))
```

## Summarizing multiple columns

* **`across()`** works here too

```{r}
#| label: summarise_across_columns
summarise(metabric, across(c(FOXA1, MLPH), mean))
```

* `everything()` selects all columns

```{r}
#| label: summarise_across_everything
metabric |>
    select(ESR1:MLPH) |>
    summarise(across(everything(), mean))
```

* **Gotcha:** the function must make sense for every column -- expect warnings and `NA` for character columns

```{r}
#| label: summarise_across_na_rm
metabric |>
    summarise(across(everything(), ~mean(.x, na.rm = TRUE)))
```

* Fix with **`where()`** plus a logical test such as `is.numeric`

```{r}
#| label: summarise_across_where_numeric
metabric |>
    summarise(across(where(is.numeric), ~mean(.x, na.rm = TRUE)))
```

* More than one function -- pass a **list**

```{r}
#| label: summarise_across_multiple_functions
metabric |>
    summarise(across(c(ESR1, ERBB2, PGR), list(mean, sd)))
```

* Name the list elements to control the output column names

```{r}
#| label: summarise_across_named_functions
metabric |>
    summarise(across(c(ESR1, ERBB2, PGR), list(Average = mean, stdev = sd)))
```

* Anonymous functions work in `summarise()` too -- correlate every gene against FOXA1
* Note the `-FOXA1` to exclude it from the selection (correlation with itself = 1)

```{r}
#| label: summarise_across_correlations
metabric |>
    summarise(across(c(ESR1:MLPH, -FOXA1), ~cor(.x, FOXA1)))
```

---

# Faceting with ggplot2

* **Faceting** splits a plot into subplots, each showing a subset of the data, by one or more categorical variables
* Two functions: **`facet_wrap()`** and **`facet_grid()`**
* Start with the colour-aesthetic version from last session

```{r}
#| label: scatter_plot_2
ggplot(data = metabric,
       mapping = aes(x = GATA3, y = ESR1, colour = PR_status)) +
    geom_point(size = 0.5, alpha = 0.5)
```

* Same data, faceted instead -- one panel per category with a banner label

```{r}
#| label: scatter_plot_3
ggplot(data = metabric, mapping = aes(x = GATA3, y = ESR1)) +
    geom_point(size = 0.5, alpha = 0.5) +
    facet_wrap(vars(PR_status))
```

* Faceting variables go inside **`vars()`** -- unquoted column names, as in `select()` and `across()`

```{r}
#| label: scatter_plot_4
ggplot(data = metabric,
       mapping = aes(x = GATA3, y = ESR1, colour = PR_status)) +
    geom_point(size = 0.5, alpha = 0.5) +
    facet_wrap(vars(PR_status))
```

* Faceting beats colours once there are more than two or three groups -- e.g. the 3-gene classifier

```{r}
#| label: scatter_plot_5
ggplot(data = metabric,
       mapping = aes(x = GATA3, y = ESR1, colour = `3-gene_classifier`)) +
    geom_point(size = 0.5, alpha = 0.5) +
    facet_wrap(vars(`3-gene_classifier`))
```

* Hence the name -- panels **wrap** onto a new row; control with `nrow` / `ncol`

```{r}
#| label: scatter_plot_6
ggplot(data = metabric,
       mapping = aes(x = GATA3, y = ESR1, colour = `3-gene_classifier`)) +
    geom_point(size = 0.5, alpha = 0.5) +
    facet_wrap(vars(`3-gene_classifier`), nrow = 1)
```

```{r}
#| label: scatter_plot_7
ggplot(data = metabric,
       mapping = aes(x = GATA3, y = ESR1, colour = `3-gene_classifier`)) +
    geom_point(size = 0.5, alpha = 0.5) +
    facet_wrap(vars(`3-gene_classifier`), ncol = 2)
```

* Facet on one variable, colour by another

```{r}
#| label: scatter_plot_8
ggplot(data = metabric,
       mapping = aes(x = GATA3, y = ESR1, colour = HER2_status)) +
    geom_point(size = 0.5, alpha = 0.5) +
    facet_wrap(vars(Neoplasm_histologic_grade))
```

* Or facet on two variables -- `facet_wrap()` gives a flat ribbon of every combination

```{r}
#| label: scatter_plot_9
ggplot(data = metabric, mapping = aes(x = GATA3, y = ESR1)) +
    geom_point(size = 0.5, alpha = 0.5) +
    facet_wrap(vars(Neoplasm_histologic_grade, HER2_status))
```

* Two variables are better as a grid -- **`facet_grid()`**, rows then columns

```{r}
#| label: scatter_plot_10
ggplot(data = metabric, mapping = aes(x = GATA3, y = ESR1)) +
    geom_point(size = 0.5, alpha = 0.5) +
    facet_grid(vars(Neoplasm_histologic_grade), vars(HER2_status))
```

* Naming `rows =` and `cols =` explicitly is clearer

```{r}
#| label: scatter_plot_11
ggplot(data = metabric, mapping = aes(x = GATA3, y = ESR1, colour = PAM50)) +
    geom_point(size = 0.5, alpha = 0.5) +
    facet_grid(rows = vars(Neoplasm_histologic_grade),
               cols = vars(HER2_status))
```

* **`labeller`** -- replace raw category values with meaningful panel labels

```{r}
#| label: scatter_plot_12
grade_labels <- c("1" = "Grade I", "2" = "Grade II", "3" = "Grade III")
her2_status_labels <- c("Positive" = "HER2 positive",
                        "Negative" = "HER2 negative")

ggplot(data = metabric, mapping = aes(x = GATA3, y = ESR1, colour = PAM50)) +
    geom_point(size = 0.5, alpha = 0.5) +
    facet_grid(rows = vars(Neoplasm_histologic_grade),
               cols = vars(HER2_status),
               labeller = labeller(Neoplasm_histologic_grade = grade_labels,
                                   HER2_status = her2_status_labels))
```

* Essential once two variables share one side of the grid -- "Positive"/"Positive" tells you nothing

```{r}
#| label: scatter_plot_13
er_status_labels <- c("Positive" = "ER positive", "Negative" = "ER negative")
#
ggplot(data = metabric, mapping = aes(x = GATA3, y = ESR1, colour = PAM50)) +
    geom_point(size = 0.5, alpha = 0.5) +
    facet_grid(rows = vars(Neoplasm_histologic_grade),
               cols = vars(ER_status, HER2_status),
               labeller = labeller(Neoplasm_histologic_grade = grade_labels,
                                   ER_status = er_status_labels,
                                   HER2_status = her2_status_labels))
```

---

# Summary

* Building up workflows by chaining operations together using the pipe operator
* Sorting rows based on values in one or more columns
* Modifying a data frame by either adding new columns or modifying existing ones
* Summarizing the values in one or more columns
* Faceting of ggplot2 visualizations
