---
title: "Session 4 -- Visualizing tabular data with ggplot2"
subtitle: "Instructor notes"
# Wider body. grid is deep-merged with the project block in _quarto.yml,
# so sidebar-width, margin-width and gutter-width are inherited.
# page-layout: full would NOT widen this page: with toc-location: left
# the TOC is #quarto-sidebar, so Quarto picks the column-page-right
# layout, whose body is still capped at $grid-body-width.
grid:
  body-width: 1400px
toc-depth: 4
toc-expand: true
# Appended to the project css rather than replacing it, so fonts.css
# still loads (Quarto treats css as a mergeable list at every level).
css: assets/css/instructor.css
include-in-header: _includes/instructor-head.html
# Keeps these pages out of docs/search.json. `navbar: search: false`
# only hides the search box - the index is written regardless.
search: false
# Rendered at 480x307 and displayed at half that by instructor.css; see
# the Figures block there for which way that trade goes and why.
fig-format: png
fig-width: 5
fig-height: 3.2
---

<!-- GENERATED by _MakeInstructorNotes.R from _instructor_prompt.md.
     Committed, but do not hand-edit below the BODY marker: the next run
     that sees a changed source page overwrites it. To change the prose,
     edit the prompt; to change the layout, edit this script. -->
<!-- instructor-notes: source=session4.qmd md5=cae0e2910dfa041a7ddcd6cf3c4ddcd0 commit=b706b28f289f4c3fcadee76dcecc404b92caf0a3 dirty=no model=opus effort=high -->

```{r}
#| label: instructor-setup
#| include: false
# width/cli.width are set project-wide in _quarto.yml and are restated
# here so that this one chunk is the whole story for a generated page -
# otherwise a later edit here silently drops them.
options(width = 80, cli.width = 80,
        tibble.print_min = 2, tibble.print_max = 2,
        # A 32-column tibble spends six lines listing the columns it
        # could not fit. One line saying how many is enough here - the
        # "1,904 x 32" header already gives the shape.
        tibble.max_extra_cols = 1)

# Base data frames have no equivalent of the tibble options, so they need
# a method. It MUST let tibbles through: they are data frames too, so S3
# dispatch would land them here, and head() would throw away the real
# row count that the tibble footer reports. kable/kableExtra/DT all
# return their own classes, so none of them come through this.
registerS3method("knit_print", "data.frame", function(x, ...) {
    if (inherits(x, "tbl_df")) return(knitr::normal_print(x))
    knitr::normal_print(utils::head(x, 2L))
}, envir = asNamespace("knitr"))
```

::: {.instructor-nav}
[Contents](instructor_index.html) &nbsp;·&nbsp; [1](instructor_session1.html) &nbsp;·&nbsp; [2](instructor_session2.html) &nbsp;·&nbsp; [3](instructor_session3.html) &nbsp;·&nbsp; **4** &nbsp;·&nbsp; [5](instructor_session5.html) &nbsp;·&nbsp; [6](instructor_session6.html) &nbsp;·&nbsp; [7](instructor_session7.html) &nbsp;·&nbsp; [8](instructor_session8.html) &nbsp;·&nbsp; [Student page](session4.html)
:::

<!-- BODY -->
::: {.callout-note .objectives title="Learning objectives"}
* Understand ggplot2's grammar of graphics
* Apply the grammar to create a variety of plots including scatter plots, box plots and histograms
* Learn about how R handles categorical data
:::

# Visualizing data with ggplot2

## METABRIC data set

* Extended METABRIC -- clinical columns plus mRNA expression for ESR1, PGR, GATA3, FOXA1
* Samples without expression data removed, so fewer rows than session 3
* Spaces in column names replaced with underscores
* `3-gene_classifier` is **not a valid R name** -- starts with a number, contains a hyphen. Needs backticks: ``metabric$`3-gene_classifier` ``
* Valid names: letters, numbers, `.` and `_`; must not start with a number

```{r}
#| label: load_metabric
#| message: false
library(tidyverse)
metabric <- read_csv("data/metabric_clinical_and_expression_data.csv")
metabric
```

# Our first ggplot - a scatter plot

* Same kind of plot as base `plot()` in earlier sessions, now the ggplot2 way

```{r}
#| label: scatter_plot_1
ggplot(data = metabric) +
    geom_point(mapping = aes(x = GATA3, y = ESR1))
```

Three things specified:

1. The **data** -- a data frame (or tibble)
2. The type of plot -- a **geom**
3. Mapping of variables to visual properties -- **aesthetics**

* `x` and `y` are aesthetics too -- counter-intuitive, say it out loud
* Other aesthetics: size, shape, colour, transparency

# Aesthetic mappings

```{r}
#| label: geom_point_help
#| eval: false
?geom_point
```

* In the Aesthetics section, **bold** = required (`x`, `y`); the rest optional

* Map a variable to `colour` exactly as for `x`/`y` -- categorical first

```{r}
#| label: scatter_plot_2
ggplot(data = metabric) +
    geom_point(mapping = aes(x = GATA3, y = ESR1, colour = ER_status))
```

* Now a continuous variable

```{r}
#| label: scatter_plot_3
ggplot(data = metabric) +
    geom_point(mapping = aes(x = GATA3, y = ESR1, colour = Nottingham_prognostic_index))
```

* **Continuous -> colour scale; discrete -> discrete colours**

* Deliberate bad idea: `size` mapped to a categorical variable

```{r}
#| label: scatter_plot_4
ggplot(data = metabric) +
    geom_point(mapping = aes(x = GATA3, y = ESR1, size = ER_status))
```

* ggplot warns -- `size` is for continuous variables

* **Setting vs mapping**: to fix size for all ~2000 points, set it *outside* `aes()`

```{r}
#| label: scatter_plot_5
ggplot(data = metabric) +
    geom_point(mapping = aes(x = GATA3,
                             y = ESR1,
                             colour = Nottingham_prognostic_index),
               size = 0.5)
```

* Converse restriction -- `shape` only takes categorical

```{r}
#| label: scatter_plot_shape_error
#| eval: false
ggplot(data = metabric) +
    geom_point(mapping = aes(x = GATA3, y = ESR1, shape = Survival_time))
```

```
Error: A continuous variable can not be mapped to shape
```

```{r}
#| label: scatter_plot_6
ggplot(data = metabric) +
    geom_point(mapping = aes(x = GATA3, y = ESR1, shape = `3-gene_classifier`),
               size = 0.9)
```

* Unclassified samples have `NA` -- ggplot silently **drops missing values** (warning in console)

* Colour + shape together -- two categorical variables at once

```{r}
#| label: scatter_plot_7
ggplot(data = metabric) +
    geom_point(mapping = aes(x = GATA3,
                             y = ESR1,
                             colour = PAM50,
                             shape = `3-gene_classifier`),
              size = 0.9)
```

* `alpha` (transparency) reveals overplotting -- like `size`, set not mapped

```{r}
#| label: scatter_plot_8
ggplot(data = metabric) +
    geom_point(mapping = aes(x = GATA3, y = ESR1, colour = `3-gene_classifier`),
               size = 0.75,
               alpha = 0.5)
```

* Colours used far more than shape/size/transparency; customization of everything comes later in the course

# ggplot2 grammar

* "gg" = **grammar of graphics**

::: {.rmdblock}
**ggplot2 grammar of graphics**

**`ggplot(data = <DATA>) + <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))`**

`ggplot()` creates the plot object.

Layers can be added by adding geoms such as `geom_point()` using the `+`
operator that ggplot2 has overridden for this purpose.
:::

# Another plot type - bar chart

* Full geom list: Packages panel -> ggplot2 -> right-click -> "Show Help" -> scroll to `geom_...`
* METABRIC's *integrative clusters* -- new subtypes defined by the study
* `geom_bar` needs only `x`; it **counts** rows for you

```{r}
#| label: bar_plot_1
ggplot(data = metabric) +
    geom_bar(mapping = aes(x = Integrative_cluster))
```

* Try `colour` -- get the class to predict

```{r}
#| label: bar_plot_2
ggplot(data = metabric) +
    geom_bar(mapping = aes(x = Integrative_cluster,
                           colour = Integrative_cluster))
```

* `colour` = outline, **`fill`** = interior -- the point of this pair

```{r}
#| label: bar_plot_3
ggplot(data = metabric) +
    geom_bar(mapping = aes(x = Integrative_cluster, fill = Integrative_cluster))
```

* Fill with a *different* variable -> **stacked bar plot**

```{r}
#| label: bar_plot_4
ggplot(data = metabric) +
    geom_bar(mapping = aes(x = Integrative_cluster, fill = ER_status))
```

* Same grammar as the scatter plot -- only the geom changed

```{r}
#| label: bar_plot_5
ggplot(data = metabric) +
    geom_bar(mapping = aes(x = Integrative_cluster, fill = `3-gene_classifier`))
```

* Classic mistake -- setting a fixed colour *inside* `aes()`

```{r}
#| label: bar_plot_6
ggplot(data = metabric) +
    geom_bar(mapping = aes(x = Integrative_cluster, fill = "blue"))
```

* Inside `aes()` R invents a variable whose value is the string "blue" for every row, then assigns it a colour from the palette

```{r}
#| label: bar_plot_7
ggplot(data = metabric) +
    geom_bar(mapping = aes(x = Integrative_cluster), fill = "blue")
```

# Multiple layers

```
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

* **`+`** is overridden by ggplot2 to add layers (geoms) and components (scales, themes)
* Multiple geoms = multiple **layers**

```{r}
#| label: scatter_line_plot_1
ggplot(data = metabric) +
    geom_point(mapping = aes(x = GATA3, y = ESR1)) +
    geom_smooth(mapping = aes(x = GATA3, y = ESR1))
```

* `geom_smooth()` -- fitted trend line; shaded band is the **standard error** on the fit
* **Gotcha:** `+` must be at the *end* of the line, not the start of the next

* Duplicated mappings -- lift them into `ggplot()`

```{r}
#| label: scatter_line_plot_2
ggplot(data = metabric, mapping = aes(x = GATA3, y = ESR1)) +
    geom_point() +
    geom_smooth()
```

* `size`/`alpha` are constants and should apply only to the points -- set inside `geom_point()`

```{r}
#| label: scatter_line_plot_3
ggplot(data = metabric, mapping = aes(x = GATA3, y = ESR1)) +
    geom_point(size = 0.5, alpha = 0.5) +
    geom_smooth()
```

* Aesthetics in `ggplot()` are **global** -- inherited by every geom

```{r}
#| label: scatter_line_plot_4
ggplot(data = metabric,
       mapping = aes(x = GATA3, y = ESR1, colour = ER_status)) +
    geom_point(size = 0.5, alpha = 0.5) +
    geom_smooth()
```

* Unexpected result: two smooth lines, one per ER group -- because `colour` was inherited

```{r}
#| label: scatter_line_plot_5
ggplot(data = metabric, mapping = aes(x = GATA3, y = ESR1)) +
    geom_point(mapping = aes(colour = ER_status), size = 0.5, alpha = 0.5) +
    geom_smooth()
```

* `inherit.aes = FALSE` -- opt a layer out of the global mappings entirely

```{r}
#| label: scatter_line_plot_6
ggplot(data = metabric,
       mapping = aes(x = GATA3, y = ESR1, colour = ER_status)) +
    geom_point(size = 0.5, alpha = 0.5) +
    geom_smooth(mapping = aes(x = GATA3, y = ESR1), inherit.aes = FALSE)
```

# Box plots

```{r}
#| label: box_plot_1
ggplot(data = metabric, mapping = aes(x = ER_status, y = GATA3)) +
    geom_boxplot()
```

* Check `?geom_boxplot` for how box, whiskers and **outliers** are defined

* Add the raw points as a second layer

```{r}
#| label: box_plot_2
ggplot(data = metabric, mapping = aes(x = ER_status, y = GATA3)) +
    geom_boxplot() +
    geom_point()
```

* Points stack in a line -- `?geom_point` points to `geom_jitter` for a categorical x

```{r}
#| label: set_seed
#| include: false
set.seed(101010)
```

```{r}
#| label: box_plot_3
ggplot(data = metabric, mapping = aes(x = ER_status, y = GATA3)) +
    geom_boxplot() +
    geom_jitter()
```

* Tame it -- `width` controls the spread

```{r}
#| label: box_plot_4
ggplot(data = metabric, mapping = aes(x = ER_status, y = GATA3)) +
    geom_boxplot() +
    geom_jitter(width = 0.3, size = 0.5, alpha = 0.25)
```

* Showing points matters most with **few** observations, where a box plot hides small n; here there are too many

* `colour` on a box plot splits each box into sub-groups

```{r}
#| label: box_plot_5
ggplot(data = metabric,
       mapping = aes(x = ER_status, y = GATA3, colour = HER2_status)) +
    geom_boxplot()
```

# Time series data

* **Line plots** for time series -- METABRIC has none, so switch data set

## COVID-19 data set

* UK Government COVID-19 press conference data, transport use before/after lockdown
* Reading from Excel needs care -- multiple sheets, header lines and footnotes
* Date column has no heading, so `skip` it and supply `col_names`; `col_types` gives read_excel a heads-up

```{r}
#| label: load_transport_use
#| message: false
#| warning: false
library(readxl)
transport_use <- read_excel("data/2020-04-26_COVID-19_Press_Conference_Data.xlsx",
                            sheet = "Transport use",
                            col_names = c("date", "transport_type", "percentage"),
                            col_types = c("date", "text", "numeric"),
                            skip = 4,
                            n_max = 186)
transport_use$percentage <- transport_use$percentage * 100
transport_use
```

* One row = traffic volume for one transport mode on one date, as a % of the first week of February

```{r}
#| label: transport_type_counts
table(transport_use$transport_type)
```

```{r}
#| label: subset_rail_use
rail_use <- transport_use[transport_use$transport_type == "National rail", ]
rail_use
```

## Dates

```{r}
#| label: date_type_and_class
typeof(rail_use$date)
class(rail_use$date)
```

* Stored as a `double` -- so it works as a continuous x aesthetic
* Class **`POSIXct`** gives it date behaviour (POSIX = cross-platform standard, `ct` = calendar time)
* Underlying number = **seconds since 1 January 1970**

```{r}
#| label: date_as_double
rail_use$date[1]
as.double(rail_use$date[1])
```

## Line plot

* **`geom_line`** -- points on successive dates joined by a line; needs `x` and `y`

```{r}
#| label: line_plot_1
ggplot(data = rail_use) +
    geom_line(mapping = aes(x = date, y = percentage))
```

* Back to the full table -- `colour` gives one line per transport mode

```{r}
#| label: line_plot_2
ggplot(data = transport_use) +
    geom_line(mapping = aes(x = date, y = percentage, colour = transport_type))
```

* `linetype` -- another aesthetic specific to lines

```{r}
#| label: line_plot_3
ggplot(data = transport_use) +
    geom_line(mapping = aes(x = date, y = percentage, linetype = transport_type))
```

# Histograms

```{r}
#| label: histogram_plot_1
ggplot(data = metabric) +
    geom_histogram(mapping = aes(x = Age_at_diagnosis))
```

* Default 30 bins triggers a warning -- set `binwidth`

```{r}
#| label: histogram_plot_2
ggplot(data = metabric) +
    geom_histogram(mapping = aes(x = Age_at_diagnosis), binwidth = 5)
```

* Or `bins` for the number of bins instead

```{r}
#| label: histogram_plot_3
ggplot(data = metabric) +
    geom_histogram(mapping = aes(x = Age_at_diagnosis), bins = 20)
```

* `colour` = bar outline, `fill` = bar interior (as with `geom_bar`)

```{r}
#| label: histogram_plot_4
ggplot(data = metabric) +
    geom_histogram(mapping = aes(x = Age_at_diagnosis), bins = 20, colour = "darkblue", fill = "grey")
```

# Categorical variables -- factors

* Categorical variables come in as characters (3-gene classifier), numbers (tumour stage) or binary (ER status)
* ggplot2 already treats **character** columns as categorical
* Problem case: a categorical variable read in as a `double`

```{r}
#| label: scatter_plot_9
ggplot(data = metabric) +
    geom_point(mapping = aes(x = GATA3, y = ESR1, colour = Tumour_stage))
```

```{r}
#| label: tumour_stage_counts
table(metabric$Tumour_stage)
```

* Only 5 discrete states but ggplot2 used a continuous colour scale -- we must tell R it is categorical
* R's type for categorical data is a **factor**

```{r}
#| label: tumour_stage_as_factor
metabric$Tumour_stage <- as.factor(metabric$Tumour_stage)
metabric[, c("Patient_ID", "Tumour_stage")]
```

* A factor is stored as **integers** plus metadata mapping each integer to a **level**

```{r}
#| label: tumour_stage_factor_properties
typeof(metabric$Tumour_stage)
class(metabric$Tumour_stage)
levels(metabric$Tumour_stage)
```

```{r}
#| label: scatter_plot_10
ggplot(data = metabric) +
    geom_point(mapping = aes(x = GATA3, y = ESR1, colour = Tumour_stage))
```

* **Level order controls plot order.** Look back at the bar plot -- 10 sat between 1 and 2, because it's a character vector (the '4ER-'/'4ER+' clusters force this) sorted alphanumerically

```{r}
#| label: integrative_cluster_as_factor
metabric$Integrative_cluster <- as.factor(metabric$Integrative_cluster)
levels(metabric$Integrative_cluster)
```

* `factor()` with the `levels` argument sets the order explicitly

```{r}
#| label: integrative_cluster_set_levels
metabric$Integrative_cluster <- factor(metabric$Integrative_cluster, 
                                       levels = c("1", "2", "3", "4ER-", 
                                                  "4ER+", "5", "6", "7", 
                                                  "8", "9", "10"))
levels(metabric$Integrative_cluster)
```

* **`forcats`** (part of the tidyverse) has functions for manipulating levels -- `fct_relevel` moves "10" to the end with far less typing

```{r}
#| label: integrative_cluster_fct_relevel
metabric$Integrative_cluster <- fct_relevel(metabric$Integrative_cluster, 
                                            "10",
                                            after = Inf)
levels(metabric$Integrative_cluster)
```

```{r}
#| label: bar_plot_8
ggplot(data = metabric) +
    geom_bar(mapping = aes(x = Integrative_cluster, fill = Integrative_cluster))
```

---

# Summary

In this session we have covered the following concepts:

* How to use ggplot2 to create different types of plot using a common grammar
* How to map variables to aesthetic features of a plot
* How to construct a plot with multiple layers (geoms)
* Various types of plot including scatter plots, box plots, bar plots and histograms
* Working with dates
* Plotting time series data as line graphs
* How to work with time series data
* Categorical variables and how R handles these with its factor type
