---
title: "Session 3 - Working with data in R"
subtitle: "Instructor notes"
# Wider body. grid is deep-merged with the project block in _quarto.yml,
# so sidebar-width, margin-width and gutter-width are inherited.
# page-layout: full would NOT widen this page: with toc-location: left
# the TOC is #quarto-sidebar, so Quarto picks the column-page-right
# layout, whose body is still capped at $grid-body-width.
grid:
  body-width: 1400px
toc-depth: 4
toc-expand: true
# Appended to the project css rather than replacing it, so fonts.css
# still loads (Quarto treats css as a mergeable list at every level).
css: assets/css/instructor.css
include-in-header: _includes/instructor-head.html
# Keeps these pages out of docs/search.json. `navbar: search: false`
# only hides the search box - the index is written regardless.
search: false
# Rendered at 480x307 and displayed at half that by instructor.css; see
# the Figures block there for which way that trade goes and why.
fig-format: png
fig-width: 5
fig-height: 3.2
---

<!-- GENERATED by _MakeInstructorNotes.R from _instructor_prompt.md.
     Committed, but do not hand-edit below the BODY marker: the next run
     that sees a changed source page overwrites it. To change the prose,
     edit the prompt; to change the layout, edit this script. -->
<!-- instructor-notes: source=session3.qmd md5=0de428a718207995911083996cb58e22 commit=25471859b6eac7133160cbe89f84a78b0ffdef91 dirty=no model=opus effort=high -->

```{r}
#| label: instructor-setup
#| include: false
# width/cli.width are set project-wide in _quarto.yml and are restated
# here so that this one chunk is the whole story for a generated page -
# otherwise a later edit here silently drops them.
options(width = 80, cli.width = 80,
        tibble.print_min = 2, tibble.print_max = 2,
        # A 32-column tibble spends six lines listing the columns it
        # could not fit. One line saying how many is enough here - the
        # "1,904 x 32" header already gives the shape.
        tibble.max_extra_cols = 1)

# Base data frames have no equivalent of the tibble options, so they need
# a method. It MUST let tibbles through: they are data frames too, so S3
# dispatch would land them here, and head() would throw away the real
# row count that the tibble footer reports. kable/kableExtra/DT all
# return their own classes, so none of them come through this.
registerS3method("knit_print", "data.frame", function(x, ...) {
    if (inherits(x, "tbl_df")) return(knitr::normal_print(x))
    knitr::normal_print(utils::head(x, 2L))
}, envir = asNamespace("knitr"))
```

::: {.instructor-nav}
[Contents](instructor_index.html) &nbsp;·&nbsp; [1](instructor_session1.html) &nbsp;·&nbsp; [2](instructor_session2.html) &nbsp;·&nbsp; **3** &nbsp;·&nbsp; [4](instructor_session4.html) &nbsp;·&nbsp; [5](instructor_session5.html) &nbsp;·&nbsp; [6](instructor_session6.html) &nbsp;·&nbsp; [7](instructor_session7.html) &nbsp;·&nbsp; [8](instructor_session8.html) &nbsp;·&nbsp; [Student page](session3.html)
:::

<!-- BODY -->
::: {.callout-note .objectives title="Learning objectives"}
* Learn how to install and use packages that extend R's basic functionality
* Read data in tabular format into R
* Calculate summary statistics on your tabular data
* Introduce the **`tibble`**, arguably the most important data structure you will use in R
* Learn how R deals with missing values
* Learn the two key **dplyr** functions for manipulating your data
    * **`filter()`** for selecting observations based on their values, i.e. selecting rows in your table
    * **`select()`** for selecting a subset of variables, i.e. selecting columns in your table
    * **`arrange()`** for sorting the observations in your table
:::

# Getting started with data in R

* **tidyverse** -- a coherent collection of packages sharing a common design philosophy, not just a bag of handy functions
* Contrast with cryptic 'base R' equivalents
* Key structures today: the **data frame** and its tidyverse derivative, the **tibble**

# Packages

* Thousands of packages extend R's core functionality
* **`library()`** loads a package -- must be done every session before its functions are available
* `tibble` is not loaded at startup

```{r}
#| label: library_tibble
#| eval: false
library(tibble)
```

* Easier to load the whole core tidyverse in one go

```{r}
#| label: library_tidyverse
library(tidyverse)
```

* Read the output aloud: which packages loaded (`tibble`, `ggplot2` -- next session)
* **Conflicts** warning -- `dplyr` functions **mask** same-named `stats` functions; use `stats::filter()` to reach the masked one

## Installing packages

* Show this error -- it means "not installed", not "typo"

```
> library(tidyverse)
Error in library(tidyverse) : there is no package called ‘tidyverse’
```

```{r}
#| label: install_tidyverse
#| eval: false
install.packages("tidyverse")
```

* Install once, `library()` every session
* At a plain terminal you may be asked for a CRAN mirror
* Binary vs source prompt -- choose **binary** (source needs a compiler)

# Tibbles

* A tibble is a data frame with the quirky, annoying behaviours fixed

## Creating a tibble

```{r}
#| label: tibble_create
wildlife <- tibble(
    animal = c("lion", "elephant", "crocodile", "hippo"),
    weight_kg = c(190, 4000, 500, 1500),
    diet = c("meat", "plants", "meat", "plants")
)
wildlife
```

* Converting an existing data frame

```{r}
#| label: tibble_from_data_frame
irisTibble <- as_tibble(iris)
```

* Still a data frame -- it has just gained the classes `tbl_df` and `tbl`

```{r}
#| label: tibble_class
class(iris)
class(irisTibble)
```

## Tibbles vs data frames

* Prints data type under each column name
* Prints only first 10 rows and only columns that fit the screen -- use `print()` for more
* Because printing already tells you structure, types and size, `str()` is much less needed

```{r}
#| label: tibble_structure
str(irisTibble)
```

* `tbl_df`, `tbl`, `data.frame` -- these are the classes a tibble **inherits** from, so most data frame functions still work

# Reading data into R

* `readr` (part of tidyverse) reads tabular files into tibbles and writes them out
* METABRIC breast cancer clinical data from cBioPortal; 2509 primary tumours, 173-gene panel
* Files are in the **data** directory inside **R_Intro**
* ".tsv" = tab-separated, so use **`read_tsv()`**

```{r}
#| label: read_metabric_tsv
metabric <- read_tsv("data/brca_metabric_clinical_data.tsv")
```

* Message reports the **column types** it guessed -- only the first few here

```{r}
#| label: print_metabric
metabric
```

::: {.rmdblock}
**`read_csv()`** and **`read_tsv()`**

* `read_csv()` -- comma-separated files; `read_tsv()` -- tab-delimited
* Both expect column names on line 1 and guess column types
* Arguments let you skip comment lines (e.g. "#") or set column types explicitly
:::

* Point out the **back ticks** around column names -- spaces, dashes, one starting with a number
* R only allows "_" and "." in names, and no leading number
* `clean_names()` from `janitor` fixes them all at once

```{r}
#| label: clean_names
#install.packages("janitor")
library(janitor)
metabric <- clean_names(metabric)
```

* **`table()`** counts occurrences of each value

```{r}
#| label: table_chemotherapy
table(metabric$chemotherapy)
```

* `summary()` on a numeric column

```{r}
#| label: summary_mutation_count
summary(metabric$mutation_count)
```

# Missing values

* Two missing mutation counts visible in the first 10 rows

```{r}
#| label: missing_values_mutation_count
metabric$mutation_count[1:10]
```

::: {.rmdblock}
**Missing values** (**`NA`**)

* `NA` = 'not available'
:::

# Data manipulation with dplyr

* **dplyr** -- loaded with the tidyverse; functions for filtering, modifying and summarising tabular data
* Switch now to the combined clinical + expression data set

```{r}
#| label: read_metabric_csv
#| message: false
metabric <- read_csv("data/metabric_clinical_and_expression_data.csv")
metabric
```

# Data semantics

* **Observations are rows**, **variables are columns** -- say it explicitly

::: {.rmdblock}
From **"Tidy Data"** by **Hadley Wickham**, [*The Journal of Statistical
Software*, vol. 59, 2014](https://www.jstatsoft.org/index.php/jss/article/view/v059i10/v59i10.pdf).

* A data set is a collection of **values**; every value belongs to a variable and an observation
* A **variable** -- all values measuring the same attribute across units
* An **observation** -- all values measured on the same unit across attributes
:::

# dplyr verbs

* **`filter()`** for filtering rows
* **`select()`** for selecting columns
* **`arrange()`** for sorting rows
* **`mutate()`** for modifying columns or creating new ones
* **`summarise()`** for computing summary values

* `group_by()` combines with these to work per group -- covered later
* Called **"verbs"** in a grammar of data manipulation; shared syntax:
    * First argument is a data frame (or tibble)
    * Further arguments describe the operation
    * Column names unquoted
    * Result is a new data frame

# Filtering rows with `filter()`

* Picks rows matching criteria built from comparison operators `==`, `>`, `<`, `!=`

```{r}
#| label: filter_died_of_disease
deceased <- filter(metabric, Vital_status == "Died of Disease")
deceased
```

* **`==`** tests equality -- not assignment
* Equivalent to base R subsetting with a logical vector

```{r}
#| label: filter_base_r_equivalent
#| eval: false
filter(metabric, Vital_status == "Died of Disease")
# is equivalent to
metabric[metabric$Vital_status == "Died of Disease", ]
```

* dplyr version: `metabric` written once, column name unquoted

```{r}
#| label: table_vital_status
table(metabric$Vital_status)
```

* `!=` -- everything that isn't "Living"

```{r}
#| label: filter_not_living
deceased <- filter(metabric, Vital_status != "Living")
nrow(deceased)
```

* **`%in%`** -- test membership of a set of values

```{r}
#| label: filter_in_vector
deceased <- filter(metabric,
                   Vital_status %in% c("Died of Disease", "Died of Other Causes"))
nrow(deceased)
```

* **`stringr`** (tidyverse) for text; **`str_starts()`** matches the start of a string

```{r}
#| label: filter_str_starts
deceased <- filter(metabric, str_starts(Vital_status, "Died"))
nrow(deceased)
```

* `str_starts()` returns a **logical vector** -- the filter condition must be `TRUE`/`FALSE` per row
* Also `str_ends()`, and `str_detect()` for regular expressions (out of scope)
* Next: filtering on a logical variable -- we have none, so create one

```{r}
#| label: filter_logical_variable
# create a new logical variable called 'Deceased'
metabric$Deceased <- metabric$Survival_status == "DECEASED"
#
# filtering based on a logical variable - only selects TRUE values
deceased <- filter(metabric, Deceased)
#
# only display those columns we're interested in
deceased[, c("Patient_ID", "Survival_status", "Vital_status", "Deceased")]
```

* **`!`** negates

```{r}
#| label: filter_negated_logical
#| eval: false
filter(metabric, !Deceased)
```

* Row count differs by one from the `Vital_status` filters -- one patient has a missing vital status
* **`is.na()`** is the only way to test for `NA`

```{r}
#| label: filter_missing_vital_status
missing_vital_status <- filter(metabric, is.na(Vital_status))
missing_vital_status[, c("Patient_ID",
                         "Survival_status",
                         "Vital_status",
                         "Deceased")]
```

* Key gotcha: `filter()` keeps only `TRUE` -- both `FALSE` **and `NA`** are dropped
* Multiple conditions, comma separated

```{r}
#| label: filter_two_conditions
filter(metabric, Survival_status == "LIVING", Survival_time > 120)
```

```{r}
#| label: filter_two_conditions_base_r
#| eval: false
metabric[metabric$Survival_status == "LIVING" & metabric$Survival_time > 120, ]
```

* Base R gets steadily more unreadable with each extra condition
* Comma is equivalent to **`&`** (Boolean AND)

```{r}
#| label: filter_comma_is_and
#| eval: false
filter(metabric, Survival_status == "LIVING", Survival_time > 120)
```

* **`|`** (Boolean OR) for either/or; combine both for complex criteria

```{r}
#| label: filter_and_or_combined
selected_patients <- filter(metabric,
                            Tumour_stage >= 3,
                            ER_status == "Positive" | PR_status == "Positive")
nrow(selected_patients)
```

* Swap comma for `&` and you must mind **operator precedence** -- use parentheses

```{r}
#| label: filter_and_or_parentheses
#| eval: false
filter(metabric, Tumour_stage >= 3 &
           (ER_status == "Positive" | PR_status == "Positive"))
```

# Selecting columns with `select()`

* Matters here -- METABRIC has too many columns to print
* `$` fine for one column; `[]` with a character vector is clumsy for several

```{r}
#| label: select_base_r
deceased[, c("Patient_ID", "Survival_status", "Vital_status", "Deceased")]
```

```{r}
#| label: select_columns
select(metabric, Patient_ID, Survival_status, Vital_status, Deceased)
```

* Same shape as `filter()`: data frame first, unquoted column names, returns a data frame even for one column
* Argument order sets column order

```{r}
#| label: select_reorder
select(metabric, Patient_ID, Vital_status, Survival_status, Deceased)
```

* **`:`** selects a range of adjacent columns

```{r}
#| label: select_range
select(metabric, Patient_ID, Chemotherapy:Tumour_stage)
```

* Helper functions -- see `?select`

```{r}
#| label: select_contains
select(metabric, contains("status"))
```

```{r}
#| label: select_ends_with
select(metabric, ends_with("status"))
```

* **`-`** drops columns instead

```{r}
#| label: select_drop_column
select(metabric, -Cohort)
```

* Names, ranges, helpers and negation can all be mixed

```{r}
#| label: select_combined_helpers
selected_columns <- select(metabric,
                           Patient_ID,
                           starts_with("Tumour_"),
                           `3-gene_classifier`:Integrative_cluster,
                           -Cellularity)
selected_columns
```

* `everything()` -- bring a few columns to the left, keep the rest in place

```{r}
#| label: select_everything
select(metabric, Patient_ID, Survival_status, Tumour_stage, everything())
```

* `relocate()` does the same job

```{r}
#| label: relocate_columns
relocate(metabric, Patient_ID, Survival_status, Tumour_stage)
```

* Renaming while selecting -- **new name on the left**

```{r}
#| label: select_with_rename
select(metabric,
       Patient_ID, Classification = `3-gene_classifier`,
       NPI = Nottingham_prognostic_index)
```

* `rename()` when you want to rename without subsetting

```{r}
#| label: rename_column
rename(metabric, PatientID = Patient_ID)
```

# Sorting using `arrange()`

* Ascending by default

```{r}
#| label: arrange_survival_time
arrange(metabric, Survival_time)
```

* `desc()` -- a helper that works only inside `arrange()`

```{r}
#| label: arrange_descending
arrange(metabric, desc(Survival_time))
```

```{r}
#| label: arrange_multiple_columns
arrange(metabric, Tumour_stage, Nottingham_prognostic_index)
```

* Second variable breaks **ties** in the first
* Usually one of the last steps, before printing or writing out

# Writing data to files

* `write_csv()` / `write_tsv()` from `readr` -- plain text, column names on line 1, openable in Excel

```{r}
#| label: write_table
#| eval: false
write_csv(metabric, "filtered_metabric_clinical_data.csv")
write_tsv(metabric, "filtered_metabric_clinical_data.tsv")
```

# Reading data from Excel files into R

* `readxl` -- installed with the tidyverse but **not** loaded by `library(tidyverse)`; load it separately
* Main functions `read_excel`, `read_xls`, `read_xlsx`
* `?read_xls` for specifying cell ranges and worksheets
