---
title: "Exploring the Metabric data set"
subtitle: "An example of using Quarto"
author: "Ashley D. Sawle"
date: "21st August 2026"
format:
  html:
    toc: true
    toc-depth: 4
    toc-location: left
    embed-resources: true
execute:
  echo: false
  warning: false
---

```         
Code chunk: "setup"
```

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(patchwork)
library(knitr)
library(kableExtra)
```

## Preamble

The purpose of this document is to show how Quarto can be used to generate a
report in HTML format.

## The data

We'll be using an extended version of the METABRIC data set in which columns
have been added for the mRNA expression values for selected genes, including
estrogen receptor alpha (**ESR1**), progesterone receptor (**PGR**), **GATA3**
and **FOXA1**.

The METABRIC study characterized the genomic mutations and gene expression
profiles for over 2000 primary breast tumours. 

[Curtis *et al.*, Nature 486:346-52, 2012](https://pubmed.ncbi.nlm.nih.gov/22522925)

[Pereira *et al.*, Nature Communications 7:11479, 2016](https://www.ncbi.nlm.nih.gov/pubmed/27161491)

Both the clinical data and the gene expression values were downloaded from
[cBioPortal](https://www.cbioportal.org/study/summary?id=brca_metabric).

```{r}
#| label: loadData
#| echo: true
#| message: false
metabric <- read_csv("data/metabric_clinical_and_expression_data.csv") |>
    select(Patient_ID,
           ER_status,
           HER2_status,
           `3-gene_classifier`,
           PAM50, ESR1, ERBB2, PGR, TP53, PIK3CA, GATA3, FOXA1, MLPH)
```

The table below shows the first few rows of the data for the first 7 columns.

```         
Code chunk: top few rows of table. Note the alignment in the columns.
```

```{r}
#| label: shortTable
#| html-table-processing: none
metabric |>
    select(1:7) |>
    slice(1:8) |>
    kbl(align = c("l", "c", "c", "c", "c", "r", "r"), digits = 2) |>
    kable_styling(bootstrap_options = c("striped", "hover")) |>
    column_spec(column = 1, bold = TRUE, border_right = TRUE)
```

## Gene expression by ER status

The figures below show gene expression distribution separated according to ER
status. The data are further split by HER2 status.

Four genes are shown:

* ESR1
* ERBB2
* GATA3
* TP53

```         
Code chunks: 1 chunk - set colours for plotting. 4 chunks - one for each plot.
```

```{r}
#| label: ersCols
ersCols <- c(Negative = "#568513", Positive = "#bb802c")
```

::: {.panel-tabset}

### ESR1

```{r}
#| label: boxESR1
#| fig-width: 6
#| fig-height: 3
ggplot(data = metabric, mapping = aes(x = ER_status, y = ESR1)) +
  geom_boxplot(aes(fill = HER2_status)) +
  scale_fill_manual(values = ersCols) +
  labs(x = "ER status") +
  theme_minimal()
```

### ERBB2

```{r}
#| label: boxERBB2
#| fig-width: 6
#| fig-height: 3
ggplot(data = metabric, mapping = aes(x = ER_status, y = ERBB2)) +
  geom_boxplot(aes(fill = HER2_status)) +
  scale_fill_manual(values = ersCols) +
  labs(x = "ER status") +
  theme_minimal()
```

### GATA3

```{r}
#| label: boxGATA3
#| fig-width: 6
#| fig-height: 3
ggplot(data = metabric, mapping = aes(x = ER_status, y = GATA3)) +
  geom_boxplot(aes(fill = HER2_status)) +
  scale_fill_manual(values = ersCols) +
  labs(x = "ER status") +
  theme_minimal()
```

### TP53

```{r}
#| label: boxTP53
#| fig-width: 6
#| fig-height: 3
ggplot(data = metabric, mapping = aes(x = ER_status, y = TP53)) +
  geom_boxplot(aes(fill = HER2_status)) +
  scale_fill_manual(values = ersCols) +
  labs(x = "ER status") +
  theme_minimal()
```

:::

## PAM50 cancer subtype

The PAM50 classification of breast cancer included 5 basic subtypes. An additional
"claudin low" subtype is also included in metabric and some tumours are not classified (NC).
The seven categories included in metabric are:

| Category   | Description    |
| :--------- | :------------- |
|Basal       | Basal-like     |
|Her2        | HER2 enriched  |
|LumA        | Luminal A      |
|LumB        | Luminal B      |
|NC          | Not classified |
|Normal      | Normal-like    |
|claudin-low | Claudin Low    |

The table below shows the median gene expression for 8 genes across all samples
according to their PAM50 cancer subtype.

```         
Code chunk: summarise expression data and generate table
```

```{r}
#| label: medianExp
#| html-table-processing: none
metabric |>
    select(PAM50, ESR1, ERBB2, PGR, TP53, PIK3CA, GATA3, FOXA1, MLPH) |>
    summarise(across(everything(), median), .by = PAM50) |>
    mutate(across(-PAM50, ~round(.x, 2))) |>
    kbl() |>
    kable_styling(bootstrap_options = c("striped", "hover")) |>
    column_spec(column = 1, bold = TRUE, border_right = TRUE)
```

## Three gene classifier

Here we look at GATA3 and ESR1 expression according to the three gene classifier.

```         
Code chunk: 4 plots arranged using `patchwork`. Caption in chunk options.
```

```{r}
#| label: threeGC
#| fig-width: 14
#| fig-height: 7
#| fig-cap: "Expression of ESR1 and GATA3 according to 3-gene classifier. (a) Bar plot showing numbers of samples in each category. (b) Scatter plot showing ESR1 expression against GATA3 expression for each sample. (c) Violin plot showing distribution of ESR1 expression. (d) Violin plot showing distribution of GATA3 expression."
p1 <- metabric |>
    ggplot(aes(x = `3-gene_classifier`)) +
    geom_bar(aes(fill = `3-gene_classifier`), colour = "black") +
    theme_minimal() +
    labs(y = "Number of samples") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

p2 <- metabric |>
    ggplot(aes(x = GATA3, y = ESR1)) +
    geom_point(aes(fill = `3-gene_classifier`), shape = 21, size = 1.5) +
    guides(fill = "none") +
    theme_minimal()

p3 <- metabric |>
    ggplot(aes(x = `3-gene_classifier`, y = ESR1)) +
    geom_violin(aes(fill = `3-gene_classifier`), colour = "black") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1))

p4 <- metabric |>
    ggplot(aes(x = `3-gene_classifier`, y = GATA3)) +
    geom_violin(aes(fill = `3-gene_classifier`), colour = "black") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1))

p1 + p2 + (p3 / p4) + 
    plot_layout(guides = "collect") +
    plot_annotation(tag_levels = "a")
```

--------------------------------------------------------------------------------
END OF REPORT

## Session Info

```{r}
#| label: sessionInfo
#| echo: true
devtools::session_info()
```
